---
title: "More Examples of Linear Model with Interactions"
author: "F.A. Barrios"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

```{r}
library(tidyverse)
library(multcomp)
library(HSAUR2)
library(wordcloud)
```

# More examples Multiple linear model with interactions  

In general in R we can write the interaction term as the product of the regresors that we are studying the interaction: $variable_1:varible_2$ for a multiple linear model with two regressors and interaction the equation looks like:  

$$ \tt Y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 $$  

## HSAUR2 example  

This are only additive linear terms to explain a random response variable $Y$ and the adjusted parameters are the $\beta_i$ of the independent variables or predictors. These variables are random too. They can be numbers and factors. Example of multiple linear regression using the clouds data clouds from HSAUR books.  

```{r}
data(clouds)
head(clouds)

# looking the data for rainfall
boxplot(rainfall ~ seeding, data = clouds)
# boxplot(rainfall ~ echomotion, data=clouds)
layout(matrix(1:2, ncol = 2))
boxplot(rainfall ~ seeding, data = clouds, ylab = "Rainfall", xlab = "Seeding")
boxplot(rainfall ~ echomotion, data = clouds, ylab = "Rainfall", xlab = "Echo Motion")
# 
layout(matrix(1:4, nrow = 2))
plot(rainfall ~ time, data = clouds)
plot(rainfall ~ cloudcover, data = clouds)
plot(rainfall ~ sne, data = clouds, xlab="S-Ne criterion")
plot(rainfall ~ prewetness, data = clouds)

#
clouds_formula <- rainfall ~ seeding + seeding:(sne+cloudcover+prewetness+echomotion) + time
Xstar <- model.matrix(clouds_formula, data = clouds)
attr(Xstar, "contrasts")
clouds_lm <- lm(clouds_formula, data = clouds)
summary(clouds_lm)

layout(matrix(1:1, nrow = 1))
# to list the betas* with the:
betaStar <- coef(clouds_lm)
betaStar

# to understand the relation of seeding and sne
psymb <- as.numeric(clouds$seeding)
plot(rainfall ~ sne, data = clouds, pch = psymb, xlab = "S-Ne criterion")
abline(lm(rainfall ~ sne, data = clouds, subset = seeding == "no"))
abline(lm(rainfall ~ sne, data = clouds, subset = seeding == "yes"), lty = 2)
legend("topright", legend = c("No seeding", "Seeding"), pch = 1:2, lty = 1:2, bty = "n")
#
# and the Covariant matrix Cov(beta*) with:
VbetaStar <- vcov(clouds_lm)

# Where the square roots of the diagonal elements are the standart errors 
sqrt(diag(VbetaStar))
clouds_resid <- residuals(clouds_lm)
clouds_fitted <- fitted(clouds_lm)

# residuals and the fitted values can be used to construct diagnostic plot
plot(clouds_fitted, clouds_resid, xlab = "Fitted values", ylab = "Residuals", type = "n", ylim = max(abs(clouds_resid)) * c(-1, 1))
abline(h = 0, lty = 2)
textplot(clouds_fitted, clouds_resid, words = rownames(clouds), new = FALSE)
qqnorm(clouds_resid, ylab = "Residuals")
qqline(clouds_resid)
```
