---
title: "Factor Analysis"
author: "F.A. Barrios, Ph.D."
date: "2024-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Factor analysis introduction. 

Factor analysis operates on the principle that observed variables are influenced by unobserved, latent factors. The main objectives of this technique include:  
	1.	Identifying the number of underlying factors.  
	2.	Determining the relationship between variables and factors.  
	3.	Interpreting the meaning of these factors.  
	4.	Quantifying how well each variable represents a factor.  

There are two primary types of factor analysis:  
Exploratory Factor Analysis (EFA): Used to uncover the underlying structure of a large set of variables without preconceived notions about the relationships.  
Confirmatory Factor Analysis (CFA): Used to test specific hypotheses about the relationship between observed variables and underlying factors.  

```{r echo = FALSE}
# Example for factor analysis
# loading libraries

library(tidyverse)
library(psych)
library(psychTools)
library(factoextra)
```

## How to do a Factor Analysis. 

The factor analysis process typically involves the following steps:  
	1.	Data Collection: Gather a large dataset with multiple variables.  
	2.	Factor Extraction: Use methods like Maximum Likelihood (ML) or Principal Axis Factoring (PAF) to extract factors.  
	3.	Determining Factor Numbers: Use techniques like scree plots to decide how many factors to retain.  
	4.	Factor Rotation: Apply rotation methods (e.g., Varimax) to improve interpretability of factors.  
	5.	Factor Loading Analysis: Examine the strength of relationships between variables and factors.  

```{r}
wdbc <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data", col_names = FALSE, show_col_types = FALSE)

# The file comes without colume names, we will use features and qualifications.
features <- c("radius", "texture", "perimeter", "area", "smoothness", "compactness", "concavity", 
              "concave_points", "symmetry", "fractal_dimension")
names(wdbc) <- c("id", "diagnosis", paste0(features,"_mean"), paste0(features,"_se"),
                 paste0(features,"_worst"))

# The data is used at Mike's PCA example.
# We will try the factor analysis subroutines with psych package

# Now for selection only the data without the ID and group
wdbc_all <- wdbc %>% select(-1, -2)
wdbc_mean <- wdbc_all %>% select(radius_mean:fractal_dimension_mean)

# First look at the variables in a correlation matrix
pairs.panels(wdbc_mean, pch = 21, stars = TRUE,  smoother = TRUE)

# starting with latent variable exploratory factor analysis (EFA)
# the PCS for all
# principal(wdbc_all, 30, rotate = "none", scores = TRUE, cor = "cor", method = "regression")
pca_wdbc_all <- principal(wdbc_all, 30, rotate = "none", scores = TRUE, use ="pairwise", cor = "cor", covar = TRUE, method = "regression")
plot(pca_wdbc_all$values, type = "b", ylab = "Eigenvalues", xlab = "Component Number")

scores_pca_wdbc <- pca_wdbc_all$scores
plot(scores_pca_wdbc[,1], scores_pca_wdbc[,2], xlab="PC1", ylab="PC2", main="Principal Components")
text(scores_pca_wdbc[,1], scores_pca_wdbc[,2], labels=rownames(scores_pca_wdbc), cex=0.7)

pca_wdbc_4c <- principal(wdbc_all, 4, rotate = "none", scores = TRUE, cor = "cor", method = "regression")
biplot(pca_wdbc_4c)

fa_wdbc_4c <- fa(wdbc_all, 4, rotate="varimax", fm="ml", oblique.scores=TRUE)
biplot(fa_wdbc_4c)

pca_wdbc_2c <- principal(wdbc_all, 2, rotate = "none", scores = TRUE, cor = "cor", method = "regression")
biplot(pca_wdbc_2c)

pca_wdbc_2c_loadings_df <- as_tibble(pca_wdbc_2c$loadings[])
pca_wdbc_2c_loadings_df$variable <- rownames(pca_wdbc_2c_loadings_df)
pca_wdbc_2c_loadings_long <- pivot_longer(pca_wdbc_2c_loadings_df, -variable, names_to = "component", values_to = "loading")

ggplot(pca_wdbc_2c_loadings_long, aes(x = variable, y = loading, fill = component)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Principal Component Loadings", x = "Variables", y = "Loading")

fa(wdbc_mean, 2, rotate="varimax", fm="ml", oblique.scores=TRUE)
iclust(wdbc_mean, nclusters=2)

```

