---
title: "Linear equations"
author: "F.A. Barrios, Ph.D."
date: "2025-09-02"
output:
  rmdformats::robobook:
    highlight: kate
  pdf_document:
description: "to prepare Class2025 presentations"
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="80")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=80)
```

## Solving systems of linear equations

In R there is a function to solve sistems of linear equations: `solve()`. It is used to solve linear algebraic equations of the form `a %*% x = b`, with *a* the coefficient matrix, *x* the vector or matrix of unknown variables and *b* is the vector or matrix of constant terms, `solve()` takes two arguments *a* and *b*.

$$  \left \{ \begin{array}{ll}
            2x + 3y  = 10 \\
            4x - 2y  = 6
            \end{array}
    \right .
$$

```{r}
# We want to write the resuts as fractions or integer numbers.
library(fractional)

a <- matrix(c(2, 3, 4, -2), nrow = 2, byrow = "TRUE")
b <- c(10, 6)

solution <- solve(a, b)
fractional(solution)
```

## Other examples

We will solve other systems of equations with the same process.  
Example 2:  
$$  \left \{ \begin{array}{ll}
            3x + 2y - z = 7 \\
             x - y + 2z = -1 \\
            2x + 3y + 4z = 12
            \end{array}
    \right .
$$
```{r}
# First define the matrix for the equation
a2 <- matrix( c(3, 2, -1, 1, -1, 2, 2, 3, 4), nrow = 3, byrow = TRUE)
b2 <- c(7, -1, 12)
sol_exa2 <- solve(a2, b2)
sol_exa2
fractional(sol_exa2)
```

Example 3:  
$$  \left \{ \begin{array}{ll}
            x + y + z = 6 \\
            x - y + 2z = 5 \\
            x - y - 3z = 10
            \end{array}
    \right .
$$

```{r}
# for exmple 3 define the matrix for the equation
a3 <- matrix( c(1, 1, 1, 1, -1, 2, 1, -1, -3), nrow = 3, byrow = TRUE)
b3 <- c(6, 5, 10)
sol_exa3 <- solve(a3, b3)
sol_exa3
```