---
title: "Linear equations"
author: "F.A. Barrios, Ph.D."
date: "2025-09-02"
output:
  rmdformats::robobook:
    highlight: kate
  pdf_document:
description: "to prepare Class2025 presentations"
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(fractional)
## Global options
options(max.print="80")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=80)
```

# Solving systems of linear equations

In R there is a function to solve sistems of linear  equations: `solve()`. It is used to solve linear algebraic equations of 
the form `a %*% x = b`, with *a* the coefficient matrix, *x* the vector or matrix of unknown variables and *b* is the vector or 
matrix of constant terms, `solve()` takes two arguments *a* and *b*.

$$  \left \{ \begin{array}{ll}
            2x + 3y  = 10 \\
            4x - 2y  = 6
            \end{array}
    \right .
$$

```{r}

a <- matrix(c(2, 3, 4, -2), nrow = 2, byrow = "TRUE")
b <- c(10, 6)

solution <- solve(a, b)
fractional(solution)
```

## Other examples

You can also embed plots, for example:
