---
title: "Multipredictor Regression with Interaction"
author: "F.A. Barrios"
date: "<small>`r Sys.Date()`</small>"
output:
  rmdformats::robobook:
    highlight: kate
  pdf_document: default
description: "to prepare classes 2022 and homework"
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

```{r message = FALSE}
# Multilevel Linear Regression with Interaction
#
library(here)
library(tidyverse)
library(emmeans)
library(multcomp)
#
# loading the HERS data set
hers <- read_csv("DataRegressBook/Chap3/hersdata.csv")
```

# Interaction[^1]
[^1]: All notes are taken from the Regression methods in biostatistics: linear, logistic, survival, and repeated measures models, Chap 4, Vitinghoff et al.

So far, we have made the assumption that causal effect of the primary predictor was the same within strata defined by the covariates. However, this may not hold. In this section, we show how to use regression to model the resulting interaction, so that we can estimate causal effects that differ according to the level of a covariate. Interaction is also referred to as effect modification or moderation, and must be distinguished from both confounding and mediation.  

## Example: Hormone Therapy and statin use  

As an example of interaction, we examine whether the effect of HT on LDL cholesterol differs according to baseline statin use, using data from HERS. To do this, a constructed interaction variable is useful. Suppose both assignment to HT and use of statins at baseline are coded using indicator variables. Then, the product of these two variables is also an indicator, equal to one only for the subgroup of women who reported using statins at baseline and were randomly assigned to HT, and zero for everyone else. Now, consider the regression model:  
$$\tt E[LDL | x] = \beta_0 + \beta_1 HT + \beta_2 statins + \beta_3 HT:statins $$  
where HT is the indicator of assignment to HT (hormone therapy), $\tt statins$ the indicator of baseline statin use, and $ HT:statins$ the interaction term. The model 
$$ \tt LDL \sim HT + statins + HT:statins $$  
can be written like  
$$ \tt LDL \sim HT * statins $$
for the model with interactions in R.


```{r}
# Now for the group with Hormone Therapy the complete HERS cohort
# Model of cholesterol LDL and the effect of Hormone Therapy (HT) and
# Statin use model is at the one year visit (to se the HT effect)

hers <- mutate(hers, HT = factor(HT))
hers <- hers %>% mutate(HT = relevel(HT, ref = "placebo"))

# For the first year visit LDL levels at one year LDL1
LDL1_model <- lm(LDL1 ~ HT * statins, data = hers)
summary(LDL1_model)
confint(LDL1_model)

# For the table 4.15 of the book, to test for the linear combination 
# of the coefficients for the interaction to TEST the combination of 
# the coefficients, HT, statin and HT and statin interaction b1, b2, and b3.

coefeq <- matrix(data=0, nrow = 1, ncol = length(LDL1_model$coefficients))
colnames(coefeq) <- names(LDL1_model$coefficients)
#
coefeq
#
coefeq[1, "HThormone therapy"] <- 1
coefeq[1, "statinsyes"] <- 1
coefeq[1, "HThormone therapy:statinsyes"] <- 1
coefeq %*% LDL1_model$coefficients

coeftestAll <- glht(model = LDL1_model, linfct= coefeq, rhs= 0)
summary(coeftestAll)

# TEST the combination of the coefficients for HT and the statin interaction b1 and b3 the third part of table 4.15
coefeq[1, "statinsyes"] <- 0
coefeq %*% LDL1_model$coefficients
coeftest <- glht(model= LDL1_model, linfct= coefeq, rhs= 0)
summary(coeftest)
```

The full model is: 
$$ \tt E[LDL | x] = \beta_0 + \beta_1 HT + \beta_2 statins + \beta_3 HT:statins $$  
The coefficient for HT, $\hat \beta_1$, shoes that women who did not report stati use at baseline, the average LDL at one year was almost 18 mg/dL lower that placebo. And for the group of HT among statin users is $\beta_1 + \beta_3$ so $\beta_3$ the interaction term gives the difference between the HT and statin users group.  


```{r}
# Example of LDL after one year of HT and physact
# This is the table 4.16 of the book, effect of the hormone therapy
# combined with effects of physical activity at the one year visit
# now to estimate the linear combination of the coefficients

hers <- mutate(hers, physact = factor(physact, 
                               levels=c("much less active",
                                        "somewhat less active",
                                        "about as active",
                                        "somewhat more active",
                                        "much more active")))

LDL1phys_model <- lm(LDL1 ~ HT * physact, data = hers)

summary(LDL1phys_model)

# The coefficients test for interaction
# can be done with glht()

coef_LDL1phys <- matrix(data=0, nrow = 1, ncol = length(LDL1phys_model$coefficients))
colnames(coef_LDL1phys) <- names(LDL1phys_model$coefficients)

# E[LDL|x] = b0 + b1 HT + b2 physact + b3 HT:physact
# we will look at slope = b3 BMIc; coeff = 0, 0, 0, 0, 0, 1, 1, 1, 1
# to test of significance

coef_LDL1phys[1, "HThormone therapy:physactsomewhat less active"] <- 1
coef_LDL1phys[1, "HThormone therapy:physactabout as active"] <- 1
coef_LDL1phys[1, "HThormone therapy:physactsomewhat more active"] <- 1
coef_LDL1phys[1, "HThormone therapy:physactmuch more active"] <- 1
coef_LDL1phys %*% LDL1phys_model$coefficients

coef_LDL1phys

coef_LDL1phystest <- glht(model= LDL1phys_model, linfct= coef_LDL1phys, rhs= 0)

summary(coef_LDL1phystest)

```

Centering BMI about its mean value of $28.6 kg/m^2$, make the interpretation for statin use easier. We compute BMIc, statins is a parameter coded 1 fro users and 0 for nonusers. So the multipredictor general model is  
$$ \tt E[LDL|x] = \beta_0 + \beta_1 statins + \beta_2 BMIc + \beta_3 statins:BMIc + \beta_4 age + \beta_5 nonwhite + \beta_6 smoking + \beta_7 drinkany $$ 
the slope asociated with BMIc in this case is $\beta_2$, for women who do not use statins (statins is 0)  
$$ \tt E[LDL|x] = \beta_0  + \beta_2 BMIc + \beta_4 age + \beta_5 nonwhite + \beta_6 smoking + \beta_7 drinkany $$ 
and statins users 
$$ \tt E[LDL|x] = \beta_0 + \beta_1 + (\beta_2 + \beta_3) BMIc + \beta_4 age + \beta_5 nonwhite + \beta_6 smoking + \beta_7 drinkany $$
So for the group of women  

```{r}
# Example of BMI and statins, with interaction
# Model of LDL and BMI*statins whith age nonwhite smoking drinkany
# For the table 4.17 of the book, to test for the linear 
# combination of the coefficients for the interaction
hers <- mutate(hers, statins = factor(statins))
hers <- mutate(hers, nonwhite = factor(nonwhite))
hers <- mutate(hers, smoking = factor(smoking))
hers <- mutate(hers, drinkany = factor(drinkany))

hers <- mutate(hers, BMIc = BMI - mean(BMI, na.rm=TRUE))

LDLbmi_model <- lm(LDL ~ statins*BMIc + age + nonwhite + smoking + drinkany, data = hers)

summary(LDLbmi_model)
confint(LDLbmi_model)
```

```{r}
# Interaction model for HT effects on absolute change in LDL
# TEST the linear combination of the coefficients for LDL ~ statins*BMIc

coef_LDLbmi <- matrix(data=0, nrow = 1, ncol = length(LDLbmi_model$coefficients))
colnames(coef_LDLbmi) <- names(LDLbmi_model$coefficients)

# E[LDL|x] = b0 + b1 statins + b2 BMIc + b3 statins:BMIc + b4 age
#           + b5 nonwhite + b6 smoking + b7 drinkany
# we will look at slope = b2 + b3 BMIc; coeff = 0, 0, 1, 1, 0, 0, 0, 0

coef_LDLbmi[1, "statinsyes"] <- 0
coef_LDLbmi[1, "BMIc"] <- 1
coef_LDLbmi[1, "statinsyes:BMIc"] <- 1
coef_LDLbmi %*% LDLbmi_model$coefficients
coef_LDLbmi

coef_LDLbmitest <- glht(model= LDLbmi_model, linfct= coef_LDLbmi, rhs= 0)
summary(coef_LDLbmitest)

# Comp of regular linear model and robust and some standard error handling
# LDLbmi_model

ggplot(data = hers, mapping = aes(x = BMI, y = LDL)) + 
  labs(x = "Body Mass Index kg/m^2", 
       y = "LDL Cholesteron mg/dL", 
       title = "Interaction") + 
  geom_point(shape = 20, color = 2) + 
  geom_smooth(method = "lm", formula = y ~ x)

ggplot(data = hers, mapping = aes(x = statins, y = LDL)) + 
  labs(x="Body Mass Index kg/m^2", 
       y= "LDL Cholesteron mg/dL", 
       title = "Interaction") + 
  geom_point(shape = 20, color = 2) + 
  geom_smooth(method = "lm", formula = y ~ x)
```

## Interaction and scale  

The The effect of HT on LDL cholesterol in the HERS trial was dependent on baseline values of LDL, with larger reductions seen among women with higher baseline values. An interaction model for absolute change in LDL from baseline to the first annual visit. Note that baseline LDL is centered in this model in order to make the coefficient for hormone therapy (HT) easier to interpret.  
The coefficients in the model have the following interpretations:  

* HT: Among women with the average baseline LDL level of 135mg/dL, the effect of HT is to lower LDL an average of 15.5mg/dL over the first year of the study.  
* cLDL0: Among women assigned to placebo, each mg/dL increase in baseline LDL is associated with a 0.35mg/dL greater decrease in LDL over the first year. That is, women with higher baseline LDL experience greater decreases in the absence of treatment; this is in part due to regression to the mean and in part to greater likelihood of starting use of statins.  
* HT * cLDL0: The effect of HT is to lower LDL an additional 0.08mg/dL for each additional mg/dL in baseline LDL. In short, larger treatment effects are seen among women with higher baseline values. The interaction is statistically significant (P = 0:013).  

```{r}
# Results for Table 4.18
hers <- mutate(hers, LDLch = LDL1 - LDL)
hers <- mutate(hers, cLDL0 = LDL - mean(LDL, na.rm = TRUE))
LDLch_model <- lm(LDLch ~ HT * cLDL0, data = hers)

summary(LDLch_model)
confint(LDLch_model)

# Percent change Table 4.19
hers <- mutate(hers, LDLpctch = ((LDL1-LDL)/LDL)*100)
LDLpctch_model <- lm(LDLpctch ~ HT * cLDL0, data = hers)

summary(LDLpctch_model)
confint(LDLpctch_model)
```

# More examples Multiple linear model with interactions  

In general in R we can write the interaction term as the product of the regresors that we are studying the interaction: $variable_1:varible_2$ for a multiple linear model with two regressors and interaction the equation looks like:  

$$ \tt Y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 $$  

## HSAUR2 example  

This are only additive linear terms to explain a random response variable $Y$ and the adjusted parameters are the $\beta_i$ of the independent variables or predictors. These variables are random too. They can be numbers and factors. Example of multiple linear regression using the clouds data clouds from HSAUR books.  

```{r}
library(HSAUR2)
library(wordcloud)
data(clouds)
head(clouds)

# looking the data for rainfall
boxplot(rainfall ~ seeding, data = clouds)
# boxplot(rainfall ~ echomotion, data=clouds)
layout(matrix(1:2, ncol = 2))
boxplot(rainfall ~ seeding, data = clouds, ylab = "Rainfall", xlab = "Seeding")
boxplot(rainfall ~ echomotion, data = clouds, ylab = "Rainfall", xlab = "Echo Motion")
# 
layout(matrix(1:4, nrow = 2))
plot(rainfall ~ time, data = clouds)
plot(rainfall ~ cloudcover, data = clouds)
plot(rainfall ~ sne, data = clouds, xlab="S-Ne criterion")
plot(rainfall ~ prewetness, data = clouds)

#
clouds_formula <- rainfall ~ seeding + seeding:(sne+cloudcover+prewetness+echomotion) + time
Xstar <- model.matrix(clouds_formula, data = clouds)
attr(Xstar, "contrasts")
clouds_lm <- lm(clouds_formula, data = clouds)
summary(clouds_lm)

layout(matrix(1:1, nrow = 1))
# to list the betas* with the:
betaStar <- coef(clouds_lm)
betaStar

# to understand the relation of seeding and sne
psymb <- as.numeric(clouds$seeding)
plot(rainfall ~ sne, data = clouds, pch = psymb, xlab = "S-Ne criterion")
abline(lm(rainfall ~ sne, data = clouds, subset = seeding == "no"))
abline(lm(rainfall ~ sne, data = clouds, subset = seeding == "yes"), lty = 2)
legend("topright", legend = c("No seeding", "Seeding"), pch = 1:2, lty = 1:2, bty = "n")
#
# and the Covariant matrix Cov(beta*) with:
VbetaStar <- vcov(clouds_lm)

# Where the square roots of the diagonal elements are the standart errors 
sqrt(diag(VbetaStar))
clouds_resid <- residuals(clouds_lm)
clouds_fitted <- fitted(clouds_lm)

# residuals and the fitted values can be used to construct diagnostic plot
plot(clouds_fitted, clouds_resid, xlab = "Fitted values", ylab = "Residuals", type = "n", ylim = max(abs(clouds_resid)) * c(-1, 1))
abline(h = 0, lty = 2)
textplot(clouds_fitted, clouds_resid, words = rownames(clouds), new = FALSE)
qqnorm(clouds_resid, ylab = "Residuals")
qqline(clouds_resid)
```
