---
title: "Solution_HW_LM"
author: "F.A.Barrios"
date: "2024-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
library(tidyverse)
library(car)
library(emmeans)
```

## Solution to HW_KM

Linear models examples from the Vittinhoff book using the WCGS data base fit a multipredictor model for total cholesterol (chol) that includes the binary predictor arcus, which is coded 1 for the group with arcus senilis, a milky ring in the iris associated with high cholesterol levels, and 0 for the reference group. Save the fitted values. Now refit the model with the code for the reference group changed to 2.


```{r}
wcgs <- read_csv("https://raw.githubusercontent.com/fabarrios/Regression/master/DataRegressBook/Chap2/wcgs.csv", show_col_types = FALSE)

names(wcgs)
wcgs_n <- mutate(wcgs, arcus = factor(arcus))

wcgs_n <- wcgs_n %>% 
    mutate(smoke = smoke %>% fct_relevel("No", "Yes"))

```

## Cholesteron linear model
Multipredictor model for total cholesterol (chol) that includes the binary predictor arcus


```{r}
chol_arcus_mod1 <- lm(chol ~ arcus, data = wcgs_n)
summary(chol_arcus_mod1)

wcgs_n <- wcgs_n %>% 
  mutate(arcus = arcus %>% fct_relevel("1", "0"))

chol_arcus_mod2 <- lm(chol ~ arcus, data = wcgs_n)
summary(chol_arcus_mod2)

# start again with the original arcus states
wcgs_n <- wcgs_n %>%
  mutate(arcus = arcus %>% fct_relevel("0", "1"))

chol_mult_mod <- lm(chol ~ arcus + bmi + smoke + age, data = wcgs_n)
summary(chol_mult_mod)

# chol ~ bmi + age + smoking + arcus
wcgs_n <- wcgs_n %>% 
  mutate(arcus = arcus %>% fct_relevel("1", "0"))
chol_mult_mod_2 <- lm(chol ~ arcus + bmi + smoke + age, data = wcgs_n)
summary(chol_mult_mod_2)
```

## T-test and ANOVA
For the T-test and the ANOVA Example we estimate a T-test with the `chol ~ arcus` using the `arcus` reference level "0".
and estimate a model with `agec`

```{r}

wcgs_n <- wcgs_n %>%
  mutate(arcus = arcus %>% fct_relevel("0", "1"))
      
t.test(chol ~ arcus, data = wcgs_n)

wcgs_n <- mutate(wcgs_n, agec = factor(agec))
boxplot(chol ~ agec, data = wcgs_n)

chol_agec_mod <- lm(chol ~ agec, data = wcgs_n)
summary(chol_agec_mod)

aov_agec_mod <- aov(chol ~ agec, data = wcgs_n)
summary(aov_agec_mod)

```

         