---
title: "Linear Model Tests"
author: F.A. Barrios
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```


```{r}
library(multcomp)
library(here)
library(lmtest)
library(tidyverse)
```

## Linear model of WCGS  

The Western Collaborative Group Study (WCGS) was a large epidemiological study designed to investigate the association between the “type A” behavior pattern and coronary heart disease (CHD) (variable `behpat` : A1, A2, B1, B2).  
We will model the cholesterol level with the body mass index and compare the model with a more complex model that includes `age`, and `arcus`.  

```{r}
WCGS <- read_csv("https://raw.githubusercontent.com/fabarrios/Regression/master/DataRegressBook/Chap2/wcgs.csv",
                 show_col_types = FALSE)

# behpat personality trait
names(WCGS)
WCGS_n <- WCGS %>% mutate( arcus <- factor(arcus))
WCGS_n <- WCGS_n %>% 
    mutate(smoke = smoke %>% fct_relevel("No", "Yes"))
```

The variable `chd69` shows whether or not the subject had a coronary event.  Other interesting variable to relate to the cholesterol level in blood is the *arcus senilis*, a milky ring in the iris associated with high cholesterol levels, `arcus` is a variable with 1 for presence of the arcus senilis and 0 for the reference group.  

```{r}
# Model of the cholesterol vs BMI in middle aged men
cholBmi_mod1 <- lm(chol ~ bmi + arcus, data = WCGS_n)
summary(cholBmi_mod1)

# Plot of chol ~ bmi data
ggplot(data = WCGS_n, aes(x = bmi, y = chol, color = arcus)) + 
  labs(x = "Body Mass Index kg/m^2", 
       y = "Total Cholesteron mg/dL", 
       title = "Linear model cholesterol vs. bmi") + 
  geom_point(shape = 20, alpha = 0.3) + 
  geom_smooth(method = "lm", formula = y ~ x)

```

Other more complex model includes the `age` variable   

```{r}
# Multipredictor model2
cholBmi_mod2 <- lm(chol ~ bmi + arcus + smoke + age, data = WCGS)
summary(cholBmi_mod2)

```

## Test of two nested linear models  

We can compare two linear models, we sill use the definition of a nested model.  We have a model that we call the full model  
$$ Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \epsilon $$  
and we want to compare to a nested model.  A nested model is defined like the full model with *less* variables included in the model.  
$$ Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon $$  
To compare the two models we use the null hypotesis:  
$H_0$: both, the full and the nested models fit the data equally well. (if you cannot discard $H_0$ therefore you should take the nested model, occam's razor).  
The $H_1$: the full model significantly outperforms the nested model. (conclusion: you keep the full model).  
So, to compare `cholBmi_mod2` to `cholBmi_mod1` models with the likelihood ratio test we run.  

```{r}

# Comparing model2 (fullmodel) with model1 (nested linear model)
# likelihood ratios test
lrtest(cholBmi_mod2, cholBmi_mod1)

```

## Testing for nonlinearities 

```{r}
# Non linear testing the model between chol ~ age 
# Test of concavity/convexity (rainbow test)

chol_age_mod <- lm(chol ~ age, data = WCGS_n)
summary(chol_age_mod)
# 
raintest(chol_age_mod, fraction = 0.5, order.by = "mahalanobis", data = WCGS_n)
```

